Staged Abstract Interpreters

### 2019-03-27 paper/20_prelim.tex
mostly section 5
few typos

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetics, recursions and conditionals. Other effectful features
such as assignments can also be supported readily. In Section~\ref{cases_imp}, we
will add more imperative features to the language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguised any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursions
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetics
\end{lstlisting}

@@

Other primitive operations provide basic functionality for the interpreter.
The method @num@ and @close@ deal with primitive values, which lift literal
terms (e.g., lambdas) to our value representation (e.g., closures).
The method @get@ simply retrieves the value mapped from the identifier @x@ in
the environment and store. Conditionals and arithmetics is handled by @br0@
and @arith@, respectively. The methods @ap_clo@ takes a function value and an
argument value and then does the application. Note that the @Env@, @Store@, and
@Value@ are all wrapped by @R[_]@ since they will be known as next-stage value
when acting as compilers.

@@

\paragraph{Unstaged Monads}
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other word, we instantiate the @AnsM@ moand as a stack of @ReaderT@ and @StateT@
transformers\footnote{The question mark syntax is a kind projector
  \cite{kindprojector}, thus \texttt{StateT[IdM,Store,?]} is equivalent to \newline
  \texttt{(\{type M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is
parameterized by type @Env@, @StateT@ is parameterized by type @Store@, and the
inner-most monad @IdM@ is just an identity monad.

@@

\begin{lstlisting}
  def close(ev: EvalFun)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

@@

\begin{lstlisting}
  def ap_clo(ev: EvalFun)(rator: Value, rand: Value): Ans = rator match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → rand)
      rt <- ext_env(ev(e))(x → α)
    } yield rt
  }
\end{lstlisting}

With other primitive operations implementated, we can implement the top-level
@run@ method, where $\rho_0$ and $\sigma_0$ are initial empty environment and
store.

>>>>

We consider a call-by-value $\lambda$-calculus in direct-style, extended
with numbers, arithmetic, recursions and conditionals. Other effectful features
such as assignments can also be supported readily. In Section~\ref{cases_imp}, we
will add more imperative features to the language. Since we are mostly
interested in analyzing the dynamic behaviors of the program, we disguised any
static semantics and type system. We also assume that input programs are
well-typed and all variables are distinct. The abstract syntax is shown as
follows:

\begin{lstlisting}
  abstract class Expr
  case class Lit(i: Int) extends Expr                         // numbers
  case class Var(x: String) extends Expr                      // variables
  case class Lam(x: String, e: Expr) extends Expr             // abstractions
  case class App(e1: Expr, e2: Expr) extends Expr             // applications
  case class If0(e1: Expr, e2: Expr, e3: Expr) extends Expr   // conditionals
  case class Rec(x: String, rhs: Expr, e: Expr) extends Expr  // recursions
  case class Aop(op: String, e1: Expr, e2: Expr) extends Expr // arithmetic
\end{lstlisting}

@@

Other primitive operations provide basic functionality for the interpreter.
The method @num@ and @close@ deal with primitive values, which lift literal
terms (e.g., lambdas) to our value representation (e.g., closures).
The method @get@ simply retrieves the value mapped from the identifier @x@ in
the environment and store. Conditionals and arithmetic is handled by @br0@
and @arith@, respectively. The methods @ap_clo@ takes a function value and an
argument value and then does the application. Note that the @Env@, @Store@, and
@Value@ are all wrapped by @R[_]@ since they will be known as next-stage value
when acting as compilers.

@@

\paragraph{Unstaged Monads}
The environment and store can be modeled by reader effect and state effect,
i.e., a Reader monad and a State monad. We combine them using monad transformers
@ReaderT@ and @StateT@.
In other word, we instantiate the @AnsM@ monad as a stack of @ReaderT@ and @StateT@
transformers\footnote{The question mark syntax is a kind projector
  \cite{kindprojector}, thus \texttt{StateT[IdM,Store,?]} is equivalent to \newline
  \texttt{(\{type M[T]=StateT[IdM,Store,T]\})\#M}}, where the @ReaderT@ is
parameterized by type @Env@, @StateT@ is parameterized by type @Store@, and the
inner-most monad @IdM@ is just an identity monad.

@@

\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = CloV(λ, ρ)
\end{lstlisting}

@@

\begin{lstlisting}
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = rator match {
    case CloV(Lam(x, e), ρ: Env) => for {
      α <- alloc(x)
      _ <- set_store(α → rand)
      rt <- ext_env(ev(e))(x → α)
    } yield rt
  }
\end{lstlisting}

With other primitive operations implemented, we can implement the top-level
@run@ method, where $\rho_0$ and $\sigma_0$ are initial empty environment and
store.

### 2019-03-27 paper/30_stageconc.tex
few typos

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)]): Rep[Value]
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ + (α → v)
        ev(e)(ρ_*)(σ_*)
    }; emit_compiled_clo(f)
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def emit_ap_clo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]): Rep[(Value, Store)]
  def ap_clo(ev: EvalFun)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store
    val res: Rep[(Value, Store)] = emit_ap_clo(rator, rand, σ)
    _ <- put_store(res._2)
  } yield res._1
\end{lstlisting}

>>>>

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)]): Rep[Value]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value], Rep[Store]) => Rep[(Value, Store)] = {
      case (v: Rep[Value], σ: Rep[Store]) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ + (α → v)
        ev(e)(ρ_*)(σ_*)
    }; emit_compiled_clo(f)
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def emit_ap_clo(rator: Rep[Value], rand: Rep[Value], σ: Rep[Store]): Rep[(Value, Store)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    σ <- get_store
    val res: Rep[(Value, Store)] = emit_ap_clo(rator, rand, σ)
    _ <- put_store(res._2)
  } yield res._1
\end{lstlisting}

### 2019-03-27 paper/40_absinterp.tex
few typos

After seeing both unstaged and staged concrete interpreters, now we trun our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time abstraction, but for the purpose of clearly
illustrating the idea, we will just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components, namely
abstract values, abstract environments and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

We follow the \textit{abstracting definitional interpreters} (ADIs) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter. We first widen the concrete values to @AbsValue@: numbers
are lifted to a singleton object @IntTop@ representing the set of all numbers;
closures remain the same. Then the type @Value@ is sets of @AbsValue@ (we abuse
the type names a little bit, as the generic interface defined the abstract type
@Value@).
The address space is constrained to be finited to ensure the computatbility of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After setteling the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is importatnt that the abstract store now maps addresses to sets of abstract
values, meaning an address may point to an over-approximated set of values
occurred at runtime.

@@

To prevent the abstract interpreter from falling into nontermination, the ADI
approach uses a two-cache mechanism. Here we first provide the type
definitions, and later we will describe the algorithm in detail. A @Cache@ is a
mapping from configurations @Config@ to sets of value-store pairs, where the
configuration is a triple of current expression being evaluated, environment and
store.

@@

This oberservation leads us to the type definition of @Result@: in concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a list of such pairs
@List[(Value, Store)]@ to represent multiple results. The results is also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

@@

The value representation of lambda terms is stil @CloV@, but we lift it to a
singleton set.

\begin{lstlisting}
  def close(ev: EvalFun)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ loooks roughly the same as the concrete
interpreter. The difference is now the function @rator@ is a set of closures, so
we can not directly destruct it by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a list and lifts the elements into the
monad stack. Then we can still implement @ap_clo@ in monadic style, where the
@CloV@ comes from the @ListT@ and the nondeterminism is naturally handled.

\begin{lstlisting}
  def lift_nd[T](vs: List[T]): AnsM[T]
  def ap_clo(ev: EvalFun)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ: Env) <- lift_nd[AbsValue](f.toList)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- ext_env(ev(e))(x → α)
  } yield rt
\end{lstlisting}

@@

As we mentioned ealier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent nontermination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been widely used in abstract
interpretation or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
Rosendahl:AbsIntPL}. Since the @in@ and @out@ cache are wrapped in the monad
stack, we first defined several methods to obtain or update the @in@ and @out@
cache through the monad stack.

@@

Otherwise, we first retrieve the results from @in@ (@⊥@ if not contained in
@in@), and put it into the @out@ cache.
Then we invoke @ev@ to compute the result for this iteration.
After the evaluation, the store @σ@ may have been changed. So we update the
@out@ cache with the new result, i.e., @(v, σ)@. Similar to store update, cache
update is also point-wise join.
Note that for the recursive call of @ev@, we also instrument it by puting
@fix(ev)@ as the first argument to @ev@.

@@

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: EvalFun => EvalFun): EvalFun = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for {
            (v, s) <- lift_nd[(Value, Store)](out(cfg).toList)
             _ <- put_store(s)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

>>>>

After seeing both unstaged and staged concrete interpreters, now we turn our
focus on abstract interpreters under the same framework. We will first present a
lattice framework with binding-time abstraction, but for the purpose of clearly
illustrating the idea, we will just use simple abstract domains such as power set
lattice or point-wise map lattice. Then we show the abstract components, namely
abstract values, abstract environments and abstract stores.
The abstract interpreter we present in this section is similar to
\citet{DBLP:journals/pacmpl/DaraisLNH17}'s, it is also
context/path/flow-insensitive by our choice.  In Section \ref{cfa}, we will
instantiate a commonly-used context-sensitive analysis;
\citet{Darais:2015:GTM:2814270.2814308} shows how to achieve path- and
flow-sensitivity by varying the monads, which is also applicable to our unstaged
or staged abstract interpreter.

@@

We follow the \textit{abstracting definitional interpreters} (ADIs) approach
\cite{DBLP:journals/pacmpl/DaraisLNH17} to refactor our concrete interpreter to
an abstract interpreter. We first widen the concrete values to @AbsValue@: numbers
are lifted to a singleton object @IntTop@ representing the set of all numbers;
closures remain the same. Then the type @Value@ is sets of @AbsValue@ (we abuse
the type names a little bit, as the generic interface defined the abstract type
@Value@).
The address space is constrained to be finite to ensure the computatbility of
analysis -- for the simplest monovariant analysis, we use variable names for
addresses. After settling the @Value@ and @Addr@ type, the environments and
stores are automatically lifted to abstract environments and abstract stores,
i.e., @Map[Ident, Addr]@ and @Map[Addr, Set[AbsValue]]@ respectively.
It is important that the abstract store now maps addresses to sets of abstract
values, meaning an address may point to an over-approximated set of values
occurred at runtime.

@@

To prevent the abstract interpreter from falling into non-termination, the ADI
approach uses a two-cache mechanism. Here we first provide the type
definitions, and later we will describe the algorithm in detail. A @Cache@ is a
mapping from configurations @Config@ to sets of value-store pairs, where the
configuration is a triple of current expression being evaluated, environment and
store.

@@

This observation leads us to the type definition of @Result@: in concrete setting,
the @Result@ is a pair of @Value@ and @Store@, now we use a list of such pairs
@List[(Value, Store)]@ to represent multiple results. The results is also linked
with the resulting cache used in this iteration.
Note that the abstract binding-time annotation is wrapped to the two objects in
the @Result@ pair, but not the whole pair. Because we statically know the result
is a tuple of two elements, but we don't know how many value-store pairs we
would have, neither the content of the cache.

@@

The value representation of lambda terms is still @CloV@, but we lift it to a
singleton set.

\begin{lstlisting}
  def close(ev: Expr => Ans)(λ: Lam, ρ: Env): Value = Set(CloV(λ, ρ))
\end{lstlisting}

For function applications, @ap_clo@ looks roughly the same as the concrete
interpreter. The difference is now the function @rator@ is a set of closures, so
we can not directly destruct it by pattern matching. Instead, we use an
auxiliary function @lift_nd@ that takes a list and lifts the elements into the
monad stack. Then we can still implement @ap_clo@ in monadic style, where the
@CloV@ comes from the @ListT@ and the nondeterminism is naturally handled.

\begin{lstlisting}
  def lift_nd[T](vs: List[T]): AnsM[T]
  def ap_clo(ev: Expr => Ans)(rator: Value, rand: Value): Ans = for {
    CloV(Lam(x, e), ρ: Env) <- lift_nd[AbsValue](rator.toList)
    α <- alloc(x)
    _ <- set_store(α → rator)
    rt <- ext_env(ev(e))(x → α)
  } yield rt
\end{lstlisting}

@@

As we mentioned earlier, the ADI approach uses a two-cache mechanism to
compute the least fixed-point and prevent non-termination, which is also called a
\textit{co-inductive} caching algorithm or \textit{truncated depth-first
evaluation} \cite{Rosendahl:AbsIntPL}. It has been widely used in abstract
interpretation or fixed point computation
\cite{DBLP:journals/pacmpl/DaraisLNH17, Wei:2018:RAA:3243631.3236800,
Rosendahl:AbsIntPL}. Since the @in@ and @out@ cache are wrapped in the monad
stack, we first defined several methods to obtain or update the @in@ and @out@
cache through the monad stack.

@@

Otherwise, we first retrieve the results from @in@ (@⊥@ if not contained in
@in@), and put it into the @out@ cache.
Then we invoke @ev@ to compute the result for this iteration.
After the evaluation, the store @σ@ may have been changed. So we update the
@out@ cache with the new result, i.e., @(v, σ)@. Similar to store update, cache
update is also point-wise join.
Note that for the recursive call of @ev@, we also instrument it by putting
@fix(ev)@ as the first argument to @ev@.

@@

\begin{figure}[t!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache; val cfg = (e, ρ, σ)
    rt <- if (out.contains(cfg)) for {
            (v, s) <- lift_nd[(Value, Store)](out(cfg).toList)
             _ <- put_store(s)
          } yield v
          else for {
            _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
            v <- ev(fix(ev))(e)
            σ <- get_store
            _ <- update_out_cache(cfg, (v, σ))
          } yield v
  } yield rt
\end{lstlisting}
\vspace{-1em}
\caption{The unstaged co-inductive caching algorithm.}
\label{fig:coind_cache}
\end{figure}

### 2019-03-27 paper/50_stageabs.tex
few typos

\paragraph{Staged Monads for Abstract Interpretation}
We use the same moand stack structure as in the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For the readability, we squeeze the three
inner transformers into a single monad @RepListReaderStateM[A, B, C]@, where @A@
and @B@ are both @Cache@ when used in @AnsM@. Now the result type is a pair of
two staged value: @Rep[List[(Value, Store)]]@ and @Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type RepListReaderStateM[A, B, C] = RepListT[RepReaderT[RepStateT[RepIdM, A, ?], B, ?], C]
    type AnsM[T] = ReaderT[StateT[RepListReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[List[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(List[(Value,Store)], Cache)]): Rep[AbsValue]
  def close(ev: EvalFun)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(List[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the set of closures into the monad stack, and generate a next-stage value
using @emit_ap_clo@, which represets the result of application, and finally we
put the @out@ cache, store and values back into the monads. The @emit_ap_clo@
method takes the target closure @clo@, argument @rand@, store and caches.

\begin{lstlisting}
  def lift_nd[T](vs: Rep[List[T]]): AnsM[T]
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(List[(Value, Store)], Cache)]
  def ap_clo(ev: EvalFun)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    clo <- lift_nd[AbsValue](rator.toList)
    σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(List[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

\vspace{-1em}
\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: EvalFun => EvalFun): EvalFun = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(List[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg).toList, out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}
\vspace{-1em}

The @res@ variable is a next-stage result, which consists of a next-stage @if@
expression. The true branch simply returns a pair of queried result form the
@out@ cache and the @out@ cache. The else branch constructs a monad @m@ first,
which evaluates @e@ under the new @out@ cache. After which, we collapse the
monad @m@ with desired arguments. Finally, we obtain a placeholder @res@ since
it is a next-stage value, and put the content of @res@ back into the moand
stack.

@@

\begin{lstlisting}
  def fix_select: EvalFun = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

Our treatment to binding-times is coarse-grained: @Expr@s is static, the rest of
the world are all dynamic. But this is not always true, since the static data
has to be used somewhere with the dyanmic operation. Partial-static data is a
way to improve binding-times and optimize the generated code.

@@

\paragraph{What the generated code look like?} In the generated code, all the
primtive operations (such as @eval@, @fix@, @ap_clo@, etc.) and moandic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representions are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the use of monads internally.

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizating Abstract Abstract Machines. However, the generated abstract
bytecode still requires another small-step abstract machine to execute, which is
an additional engineering efforts.

>>>>

\paragraph{Staged Monads for Abstract Interpretation}
We use the same monad stack structure as in the unstaged abstract interpreter,
but replacing \textit{all} the transformers to their staged version. The
following code shows this change. For the readability, we squeeze the three
inner transformers into a single monad @RepListReaderStateM[A, B, C]@, where @A@
and @B@ are both @Cache@ when used in @AnsM@. Now the result type is a pair of
two staged value: @Rep[List[(Value, Store)]]@ and @Rep[Cache]@.

\begin{lstlisting}
  trait StagedAbstractSemantics extends AbstractComponents {
    type R[T] = Rep[T]
    type RepListReaderStateM[A, B, C] = RepListT[RepReaderT[RepStateT[RepIdM, A, ?], B, ?], C]
    type AnsM[T] = RepReaderT[RepStateT[RepListReaderStateM[Cache, Cache, ?], Store, ?], Env, T]
    type Result = (Rep[List[(Value, Store)]], Rep[Cache])
    ...
  }
\end{lstlisting}

@@

\begin{lstlisting}
  def emit_compiled_clo(f: (Rep[Value], Rep[Store], Rep[Cache], Rep[Cache])
                           => Rep[(List[(Value,Store)], Cache)]): Rep[AbsValue]
  def close(ev: Expr => Ans)(λ: Lam, ρ: Rep[Env]): Rep[Value] = {
    val Lam(x, e) = λ
    val f: (Rep[Value],Rep[Store],Rep[Cache],Rep[Cache]) => Rep[(List[(Value,Store)],Cache)] = {
      case (arg, σ, in, out) =>
        val α = alloc(σ, x); val ρ_* = ρ + (unit(x) → α); val σ_* = σ ⊔ Map(α → arg)
        ev(e)(ρ_*)(σ_*)(in)(out)
    }; Set[AbsValue](emit_compiled_clo(f))
  }
\end{lstlisting}

The @ap_clo@ method is also similar: we use the staged version of @lift_nd@ to
lift the set of closures into the monad stack, and generate a next-stage value
using @emit_ap_clo@, which represents the result of application, and finally we
put the @out@ cache, store and values back into the monads. The @emit_ap_clo@
method takes the target closure @clo@, argument @rand@, store and caches.

\begin{lstlisting}
  def lift_nd[T](vs: Rep[List[T]]): AnsM[T]
  def emit_ap_clo(rator: Rep[AbsValue], rand: Rep[Value], σ: Rep[Store],
                  in: Rep[Cache], out: Rep[Cache]): Rep[(List[(Value, Store)], Cache)]
  def ap_clo(ev: Expr => Ans)(rator: Rep[Value], rand: Rep[Value]): Ans = for {
    clo <- lift_nd[AbsValue](rator.toList)
    σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val res: Rep[(List[(Value, Store)], Cache)] = emit_ap_clo(clo, rand, σ, in, out)
    _ <- put_out_cache(res._2)
    vs <- lift_nd[(Value, Store)](res._1)
    _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}

@@

\vspace{-1em}
\begin{figure}[h!]
  \centering
\begin{lstlisting}
  def fix(ev: (Expr => Ans) => (Expr => Ans)): Expr => Ans = e => for {
    ρ <- ask_env; σ <- get_store; in <- ask_in_cache; out <- get_out_cache
    val cfg: Rep[Config] = (unit(e), ρ, σ)
    val res: Rep[(List[(Value, Store)], Cache)] =
      if (out.contains(cfg)) (out(cfg).toList, out) // a next-stage if expression
      else { val m: Ans = for {
               _ <- put_out_cache(out + (cfg → in.getOrElse(cfg, ⊥)))
               v <- ev(fix(ev))(e)
               σ <- get_store
               _ <- update_out_cache(cfg, (v, σ))
             } yield v
             m(ρ)(σ)(in)(out) }
    _ <- put_out_cache(res._2); vs <- lift_nd(res._1); _ <- put_store(vs._2)
  } yield vs._1
\end{lstlisting}
\vspace{-1em}
\caption{The staged co-inductive caching algorithm.}
\label{fig:staged_coind_cache}
\end{figure}
\vspace{-1em}

The @res@ variable is a next-stage result, which consists of a next-stage @if@
expression. The true branch simply returns a pair of queried result form the
@out@ cache and the @out@ cache. The else branch constructs a monad @m@ first,
which evaluates @e@ under the new @out@ cache. After which, we collapse the
monad @m@ with desired arguments. Finally, we obtain a placeholder @res@ since
it is a next-stage value, and put the content of @res@ back into the monad
stack.

@@

\begin{lstlisting}
  def fix_select: Expr => Ans = e => e match {
    case Lit(_) | Var(_) | Lam(_, _) => eval(fix_select)(e)
    case _ => fix_cache(e)
  }
\end{lstlisting}

@@

Our treatment to binding-times is coarse-grained: @Expr@s is static, the rest of
the world are all dynamic. But this is not always true, since the static data
has to be used somewhere with the dynamic operation. Partial-static data is a
way to improve binding-times and optimize the generated code.

@@

\paragraph{What the generated code look like?} In the generated code, all the
primitive operations (such as @eval@, @fix@, @ap_clo@, etc.) and monadic
operations (such as @flatMap@ and @map@) are eliminated. The residual program
consists of statements and expressions that purely manipulate the environment,
store, and two caches, whose underlying representations are all @Rep[Map[K,V]]@. We
also have several operations on tuples and lists, which are residualized from
the use of monads internally.

@@

What we implemented is a big-step, compositional abstract interpreter in monadic
style, where \textit{compositional} means that every recursive call of our abstract
interpreter is applied to proper substructures of the current syntactic
parameters \cite{10.1007/3-540-61580-6_11}. This compositionality ensures that
specialization can be done by unfolding, as well as guarantees the termination
of specialization procedure. It is also possible to specialize small-step
operational abstract semantics through abstract compilation
\cite{Boucher:1996:ACN:647473.727587} -- as
\citet{Johnson:2013:OAA:2500365.2500604} implemented it for
optimizing Abstract Abstract Machines. However, the generated abstract
byte-code still requires another small-step abstract machine to execute, which is
an additional engineering efforts.
